---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading the required packages
```{r, message=FALSE}
library(data.table)
library(ggplot2)
library(scales)
library(lubridate)
```

## Loading and preprocessing the data
Downloading and unzipping data to obtain the csv file.
```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = "Factivity.zip")
unzip("Factivity.zip")
```

Reading csv into a data.table.
```{r}
activity_data <-fread("activity.csv")
```

## What is mean total number of steps taken per day?
```{r}
total_steps_filled <- aggregate(steps ~ date, data = filled_activity_data, sum)
plot3 <- ggplot(total_steps_filled, aes(x = steps)) + 
  geom_histogram(fill = "deepskyblue4", bins = 11) + 
  labs(title = "Daily Steps with replaces NA's", x = "Step Count", y = "Frequency")
print(plot3)
```


## What is the average daily activity pattern?
Calculating the average steps taken for each 5-minute interval.
```{r}
interval_steps <- aggregate(steps ~ interval, data = activity_data, mean)
```

Converting 'interval' column data to a vailid date-time format.
```{r}
interval_steps$time <- as.character(interval_steps$interval)
for (i in 1:2){
  interval_steps$time[i] <- as.character(paste0("0",interval_steps$time[i]))
}
for (i in 1:12){
  interval_steps$time[i] <- as.character(paste0("00",interval_steps$time[i]))
}
for (i in 13:120){
  interval_steps$time[i] <- as.character(paste0("0",interval_steps$time[i]))
}
interval_steps$time <- as.POSIXct(interval_steps$time, format = "%H%M")
```

Now that we have the valid date-time format we can have a cleaner looking time-series plot, for the average 24-hour period.
```{r}
plot2 <- ggplot(interval_steps, aes(x = time, y = steps)) +
  geom_line(col = "deepskyblue4") +
  labs(title = "Time Series Plot of Average Steps Taken", x = "Time of Day", y = "Steps") +
  scale_x_datetime(labels = date_format("%H:%M", tz = "MST"), date_breaks = "4 hours")
print(plot2)
```


## Imputing missing values
The total amount of NA's and the percentage of missing step data.
```{r}
nas <- is.na(activity_data$steps)
sum(nas)
mean(nas)
```
Function to impute the mean for a 5-minute interval into the appropriate mssing data (NA's).
```{r}
replaceNas= function(steps, interval) {
  replace = NA
  if (!is.na(steps)) { 
    replace = steps } 
  else { 
    replace = interval_steps[interval_steps$interval == interval, "steps"]}
  return(replace) }
```
Apply 'replaceNas' function.
```{r}
filled_activity_data = activity_data
filled_activity_data$steps = mapply(replaceNas, filled_activity_data$steps, filled_activity_data$interval)
```


## Are there differences in activity patterns between weekdays and weekends?
Creating two new factor variables showing the day of the week, and if its a weekday of weekend. 
```{r}
filled_activity_data$day <- weekdays(filled_activity_data$date)
weekday <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
weekDayOp <- function(dayofweek) {
  fill = ""
  if (dayofweek %in% weekday) {
    fill = "Weekday" }
  else {
    fill = "Weekend" }
  return(fill) }
filled_activity_data$weekday <- mapply(weekDayOp,filled_activity_data$day)
```

Calculating the average number of steps per 5-minute interval for weekdays and weekends.
```{r}
filled_totals_day <- aggregate(steps ~ interval + weekday, data = filled_activity_data, mean)
```

Coverting 'interval' column data to a vailid date-time format.
```{r}
filled_totals_day$time <- as.character(filled_totals_day$interval)
for (i in 1:2){
  filled_totals_day$time[i] <- as.character(paste0("0",filled_totals_day$time[i]))
}
for (i in 1:12){
  filled_totals_day$time[i] <- as.character(paste0("00",filled_totals_day$time[i]))
}
for (i in 13:120){
  filled_totals_day$time[i] <- as.character(paste0("0",filled_totals_day$time[i]))
}
for (i in 289:290){
  filled_totals_day$time[i] <- as.character(paste0("0",filled_totals_day$time[i]))
}
for (i in 289:300){
  filled_totals_day$time[i] <- as.character(paste0("0",filled_totals_day$time[i]))
}
for (i in 301:408){
  filled_totals_day$time[i] <- as.character(paste0("0",filled_totals_day$time[i]))
}
filled_totals_day$time <- as.POSIXct(filled_totals_day$time, format = "%H%M")
```

Now that we have the valid date time format we can have a cleaner looking Time-Seriesplot, for comparison of the average 24-hour period on weekdays and weekends.
```{r}
plot4 <- ggplot(filled_totals_day, aes(time, steps, col = factor(weekday))) +
  facet_grid(.~factor(weekday)) +
  geom_line(show.legend = F) +
  labs(x = "Time of Day") +
  labs(y = "Steps") +
  labs(title = "Time Series Plot Comparison of Steps") +
  scale_x_datetime(labels = date_format("%H:%M", tz = "MST"), date_breaks = "6 hours")
print(plot4)
```
